---
title: "End-to-end 16S rRNA Workflow"
author: "Microbiome Analytics"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(targets)
library(tidyverse)
library(phyloseq)
```

# Overview

This document reports on an amplicon sequencing workflow implemented with **dada2**, **phyloseq**, **decontam**, **vegan**, **ANCOM-BC**, **Maaslin2**, **microbiome**, and **targets**. Run the computational pipeline with:

```r
renv::activate()
tar_make()
```

Once the pipeline completes, re-knit this R Markdown document to summarise the analysis products stored in the `targets` cache.

# Data import and quality control

```{r load-data}
tar_load(fastq_table)
tar_load(quality_profiles)
tar_load(metadata)
fastq_table
```

Quality profiles for the first few samples are saved as PNG images. Example paths:

```{r quality-profiles}
head(quality_profiles)
```

# DADA2 processing

```{r dada2}
tar_load(filtered_reads)
tar_load(error_models)
tar_load(asv_results)
tar_load(seqtab_nochim)
tar_load(contaminant_results)
filtered_reads
```

Reads retained after filtering:

```{r filter-summary}
filtered_reads |>
  select(sample_id, reads_in, reads_out)
```

Contaminant ASVs flagged by **decontam**:

```{r contaminants}
contaminant_results$removed_asvs
```

# Taxonomy, phylogeny, and phyloseq object

```{r phyloseq}
tar_load(taxonomy)
tar_load(phy_tree)
tar_load(physeq)
physeq
```

# Alpha- and beta-diversity

```{r diversity}
tar_load(alpha_diversity)
tar_load(beta_diversity)
tar_load(ordinations)
alpha_diversity
```

```{r beta}
beta_diversity$distance_tidy |>
  head()
```

Ordination coordinates joined to metadata:

```{r ordination}
tar_load(metadata)
ord_df <- combine_ord_coord(ordinations, metadata)
head(ord_df)
```

# PERMANOVA

```{r permanova}
tar_load(permanova)
permanova
```

# Differential abundance analyses

```{r da}
tar_load(ancombc_results)
tar_load(maaslin2_results)
head(ancombc_results)
```

```{r maaslin}
head(maaslin2_results)
```

# Core microbiome

```{r core}
tar_load(core_microbiome)
core_microbiome
```

# Functional predictions

```{r functional}
tar_load(function_predictions)
tar_load(pathway_analysis)
function_predictions$method
```

Top pathways differentiating groups:

```{r pathways}
pathway_analysis$summary |>
  head()
```

```{r pathway-plot}
pathway_analysis$plot
```

# Reproducibility

The workflow leverages:

- **renv** for dependency management (see `renv.lock`).
- **targets** for orchestrating the pipeline (see `_targets.R`).
- **RMarkdown** for literate documentation (this document).

To reproduce the analysis on a new machine:

1. Clone the repository.
2. Run `renv::restore()` to install required packages.
3. Place raw FASTQ files in `data/raw/` and update metadata as needed.
4. Download the SILVA/GTDB reference database into `references/` and update `_targets.R` if the paths differ.
5. Execute the pipeline with `tar_make()`.
6. Render this report using `rmarkdown::render("analysis/microbiome_workflow.Rmd")`.
